### 2022-03-07 bools/bools.tex
more

\looseness=-1
In both pure and cooperative mode, Satallax is the winner.
Zipperposition comes in close second, showing that our approach is a promising
basis for further extensions. Leo-III uses SMT solver CVC4, which features
native support for Booleans, as a backend. It is possible that the use of CVC4 is
one of the reasons for massive improvement in success rate of cooperative
configuration of Leo-III, compared with the pure version. Therefore, we conjecture
that including support for SMT backends in Zipperposition might be beneficial.

>>>>

\looseness=-1
In both pure and cooperative mode, Satallax is the winner.
Zipperposition comes in close second, showing that our approach is a promising
basis for further extensions. Leo-III uses SMT solver CVC4, which features
native support for Booleans, as a backend. It is possible that the use of CVC4 is
one of the reasons for massive improvement in success rate of cooperative
configuration of Leo-III, compared with the pure version. Therefore, we conjecture
that including support for SMT backends in Zipperposition might be beneficial.\section{Discussion and Related Work} 
\label{sect:bool:discussion}

Our work is primarily motivated by the goal of closing the gap between
higher-order ``hammer'' or software verifier frontends and first-order backends.
Considerable amount of research effort has gone into making the translations of
higher-order logic as efficient as possible. Descriptions of hammers like
HOLyHammer \cite{ku-15-holyhammer} and Sledgehammer
\cite{pb-12-sh} for Isabelle contain details of these
translations. Software verifiers Boogie \cite{lr-10-boogie} and Why3
\cite{bfcp-11-why3} use similar translations.

\looseness=-1
Established higher-order provers like Leo-III and Satallax perform very
well on TPTP benchmarks; however, recent evaluations (including the one in Sect.~\ref{sec:ehoh:evaluation}) show that on Sledgehammer
problems they are outperformed by translations to first-order logic
\cite{bbtvw-21-sup-lam, %EHOH CHAPTER,
cbetal-11-cvc4}. Those two provers are built from the ground up as
higher-order provers -- treatment of exclusively higher-order issues such as
extensionality or choice is built into them usually using explosive rules. Those
explosive rules might contribute to their suboptimal performance on mostly
first-order Sledgehammer problems.

In contrast, the approach taken in this thesis is to start with a first-order prover and gradually extend it
with higher-order features. The work performed in the context of Matryoshka project~\cite{matryoshka}, 
in which I participate, resulted in adding support for
$\lambda$-free higher-order logic with Booleans to E (Chapter~\ref{ch:ehoh})
and veriT \cite{cbetal-11-cvc4}, and adding support for Boolean-free higher-order
logic to Zipperposition. Authors of many all state-of-the-art
first-order provers have implemented some form of support for higher-order reasoning. This
is true both for SMT solvers, witnessed by the recent extension of CVC4 and veriT~\cite{cbetal-11-cvc4},
 and for superposition provers, witnessed
by the extension of Vampire \cite{br-19-restricted-unif}. All of those approaches were
arguably more focused on functional aspects of higher-order logic, such as
$\lambda$-binders and function extensionality, than on Boolean aspects such as
Boolean subterms and Boolean extensionality. A notable exception is work by
Kotelnikov et al.\ that introduced support for Boolean subterms to first-order
Vampire \cite{kotelnikov-15-fool,kotelnikov-16-fool}.

The main merit of our approach is that it combines two successful complementary
approaches to support features of higher-order logic that have not been combined
before in a modular way. It is based on a higher-order superposition calculus
that incurs around 1\% of overhead on first-order problems compared with classic
superposition \cite{bbtvw-21-sup-lam}. We conjecture that it is
this efficient reasoning base on which the approach is based that contributes to
its competitive performance.

\section{Conclusion} 
\label{sect:bool:conclusion}

We presented a pragmatic approach to support Booleans in a modern automatic prover
for clausal higher-order logic. Our approach combines
previous research efforts that extended first-order provers with complementary
features of higher-order logic. It also proposes some solutions for the issues that
emerge with this combination. The implementation shows clear improvement over previous
techniques and competitive performance.

What our work misses is an overview of heuristics that can be used to curb the
explosion incurred by some of the rules described in this paper. In the next chapter,
we explore exactly this topic.
